\ProjectName{}: Data Imputation as a Query Optimization

### 2016-12-14 paper/algorithm.tex
paper: minor word changes

For each sub-query, using the histogram information, we keep track of the estimated number of
null values in a column  ($\text{MissingCount}(c, Q)$), given the histograms associated with
the sub-query. We use this estimate to compute
$\text{Loss}(Q)$ as follows.

@@

Evaluation of the time complexity for \textit{Impute} is more thorny, as the properties of
different decision tree building algorithms can vary significantly, the underlying IO cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (yield intelligent query
plans), though the range of $\alpha$ that does promote tradeoffs is condensed.

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data}) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}. Though
we provide quantitative results for the CE-CART implementation, ImputeDB is designed to be
agnostic to the choice of imputation algorithm, and additional algorithms --- with
associated time and loss functions --- could be insert without any change in design.

@@

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (the imputed values do not change across
epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
provides details on the implementation.

The flexibility of this non-parametric imputation approach aligns well with the spirit of
query optimization, which aims to reconcile performance with a declarative interface to data
manipulation. By removing concerns for imputation, users executing queries in \ProjectName{}
don't need to consider the potential nature of the missing data.

>>>>

For each sub-query, we keep track of the estimated number of
null values in a column  ($\text{MissingCount}(c, Q)$) by using the histograms associated with
the sub-query. We use this estimate to compute
$\text{Loss}(Q)$ as follows.

@@

Evaluation of the time complexity for \textit{Impute} is thornier, as the properties of
different decision tree building algorithms can vary significantly, the underlying IO cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans), though the range of $\alpha$ that does promote tradeoffs is condensed.

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data}) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}. Though
we provide quantitative results for the CE-CART implementation, ImputeDB is designed to be
agnostic to the choice of imputation algorithm, and additional algorithms --- with
associated time and loss functions --- could be inserted without any change in design.

@@

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e. the imputed values do not change across
epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
provides details on the implementation.

The flexibility of this non-parametric imputation approach aligns well with the spirit of
query optimization, which aims to reconcile performance with a declarative interface to data
manipulation. By removing concerns for imputation, users executing queries in \ProjectName{}
don't need to consider the potential nature of missing data.

### 2016-12-14 paper/conclusion.tex
paper: minor word changes

In our experiments, we considered a series of analytic queries (using relevant aggregates) and set queries (with simple projections) on real-world
American Community Survey data
and showed that the chosen query plan improves accuracy of results relative to simply ignoring missing values. Furthermore,
the variety of imputation operator placements across queries further emphasizes the lack of flexibility imposed by the coarse-grained
pre-processing approach to imputation. By allowing different imputation strategies for different queries, we
take a fine-grained approach to missing data, better addressing the often differing needs of database users.

>>>>

In our experiments, we considered a series of analytic queries (using relevant aggregates) and set queries (with simple projections) on real-world
American Community Survey data and synthetic data
and showed that the chosen query plan improves accuracy of results relative to simply ignoring missing values. Furthermore,
the variety of imputation operator placements across queries further emphasizes the lack of flexibility imposed by the coarse-grained
pre-processing approach to imputation. By allowing different imputation strategies for different queries, we
take a fine-grained approach to missing data, better addressing the often differing needs of database users.

### 2016-12-14 paper/examples.tex
paper: Add discussion of time complexity of tree build
paper: minor word changes

The analyst submits a query (Figure~\ref{fig:example-query}) to ImputeDB which finds an optimized query
plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
emphasis parameter, in order to receive a plan that does an expensive imputation step. See
Section~\ref{sec:cost-model} for more details.}. The resulting query plan
(Figure~\ref{fig:query-plan})
imputes missing values for \verb!acs.TEL! after states with high polling error have been
selected. This decision reduces the set of tuples that must be input into the imputation
algorithm to just those that are relevant to the query. Later, we show the algorithms that
the ImputeDB query optimizer uses in order to detect these opportunities and control the
tradeoffs between efficiency and imputation quality. Finally, note that the optimizer doesn't
place any imputation operator for the \verb|acs.ST| column, as optimizer uses the histogram to
detect that there are no missing values there (as might be expected for a survey question that isn't
impacted by non-responses).

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[width=0.6\textwidth]{figures/example.png}
    \caption{Query plan generated by ImputeDB prioritizing imputation quality.}
    \label{fig:query-plan}
\end{figure}

>>>>

The analyst submits a query (Figure~\ref{fig:example-query}) to ImputeDB which finds an optimized query
plan\footnote{The analyst specifies a relatively high value of $\alpha$, the quality
emphasis parameter, in order to receive a plan that does an expensive imputation step. See
Section~\ref{sec:cost-model} for more details.}. The resulting query plan
(Figure~\ref{fig:query-plan})
imputes missing values for \verb!acs.TEL! after states with high polling error have been
selected. This decision reduces the set of tuples that must be input into the imputation
algorithm to just those that are relevant to the query. Later, we show the algorithms that
the ImputeDB query optimizer uses in order to detect these opportunities and control the
tradeoffs between efficiency and imputation quality. Finally, note that the optimizer doesn't
place any imputation operator for the \verb|acs.ST| column, as it uses a histogram to
detect that there are no missing values there (as might be expected for a survey question that isn't
impacted by non-responses).

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[width=0.6\textwidth]{figures/example.png}
    \caption{Query plan generated by ImputeDB prioritizing imputation quality}
    \label{fig:query-plan}
\end{figure}
